---
title: "3 · Import & DESeq2"
---

This is the heart of the workshop: from salmon files to a ranked list of differentially expressed genes. Drive each step by prompting Claude Code — the code below is what it should produce, and the output is real, run on this repo's data.

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

## Load the salmon output with tximport

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[Read `data/samples.csv`. Then load the four salmon `quant.sf` files with tximport, summarizing transcripts to genes using `data/tx2gene.csv`. Show me the dimensions of the gene-level count matrix.]{.prompt-text}
:::

```{r tximport}
library(tximport)
library(readr)
library(dplyr)

samples <- read_csv("data/samples.csv", show_col_types = FALSE)
samples$condition <- factor(samples$condition, levels = c("normoxia", "hypoxia"))

files <- file.path("data", "salmon", samples$sample, "quant.sf")
names(files) <- samples$sample

tx2gene <- read_csv("data/tx2gene.csv", show_col_types = FALSE)
txi <- tximport(files, type = "salmon", tx2gene = tx2gene)

dim(txi$counts)            # genes x samples
head(round(txi$counts), 3)
```

`tximport` returned three matrices — `counts`, `abundance` (TPM) and `length`. DESeq2 uses all three: the per-gene average transcript length becomes an offset that corrects for differential transcript usage between samples.

## Build the DESeq2 dataset

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[Create a DESeq2 dataset from the tximport object with design `~ condition`, and set `normoxia` as the reference level so fold changes read as hypoxia-vs-normoxia.]{.prompt-text}
:::

```{r dds}
library(DESeq2)

dds <- DESeqDataSetFromTximport(txi, colData = samples, design = ~ condition)
dds$condition <- relevel(dds$condition, ref = "normoxia")
dds
```

## Filter low-count genes

Genes with almost no reads add noise and drag down multiple-testing power. A light filter is standard:

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[Remove genes whose total count across all samples is 10 or fewer, and tell me how many genes remain.]{.prompt-text}
:::

```{r filter}
keep <- rowSums(counts(dds)) > 10
dds <- dds[keep, ]
nrow(dds)
```

## Run differential expression

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[Run DESeq2 and extract results for the contrast hypoxia vs normoxia. Summarize how many genes are up and down at padj < 0.05.]{.prompt-text}
:::

```{r deseq}
dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "hypoxia", "normoxia"))
summary(res)
```

`DESeq()` does three things in one call: estimates size factors (library-size normalization), estimates per-gene dispersion, and fits a negative-binomial GLM with a Wald test.

## Annotate with gene symbols

The results are indexed by `ENSG…` IDs. Join the symbol map so the table is human-readable:

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[Convert the results to a data frame, add gene symbols from `data/gene_name_map.csv`, sort by adjusted p-value, and show the top 15 genes.]{.prompt-text}
:::

```{r annotate}
library(tibble)

gene_name_map <- read_csv("data/gene_name_map.csv", show_col_types = FALSE)

res_df <- res |>
  as.data.frame() |>
  rownames_to_column("gene_id") |>
  left_join(gene_name_map, by = "gene_id") |>
  arrange(padj)

res_df |>
  select(gene_name, gene_id, baseMean, log2FoldChange, padj) |>
  head(15)
```

```{r sanity}
# Biological sanity check: are canonical hypoxia genes up-regulated?
hif <- c("VEGFA", "CA9", "BNIP3", "SLC2A1", "PGK1", "LDHA", "PDK1")
res_df |>
  filter(gene_name %in% hif) |>
  select(gene_name, log2FoldChange, padj) |>
  arrange(padj)
```

If the hypoxia program is up, the analysis is behaving. 🎯

```{r save}
#| include: false
# Persist for the visualization & enrichment pages.
dir.create("results", showWarnings = FALSE)
saveRDS(dds, "results/dds.rds")
write_csv(res_df, "results/de_hypoxia_vs_normoxia.csv")
```

→ Let's make this visual — **[4 · Visualization](04-visualization.qmd)**.
