---
title: "5 · Enrichment"
---

A list of significant genes is a means, not an end. Functional enrichment asks: **what biological processes are over-represented among the genes that changed?** For a hypoxia experiment, we expect terms like *response to hypoxia*, *glycolysis*, and *angiogenesis*.

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
library(dplyr); library(readr)
res_df <- read_csv("results/de_hypoxia_vs_normoxia.csv", show_col_types = FALSE)
```

## Pick the gene sets

We need two lists: the **significant** up-regulated genes, and the **universe** of all tested genes (so enrichment is measured against the right background). GENCODE IDs carry a version suffix (`.17`) that the annotation database doesn't use, so we strip it.

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[From the DESeq2 results, take genes with padj < 0.05 and log2FoldChange > 1 as the up-regulated set, and all tested genes as the universe. Strip the Ensembl version suffix from the IDs.]{.prompt-text}
:::

```{r genesets}
strip_version <- function(x) sub("\\.\\d+$", "", x)

up <- res_df |>
  filter(!is.na(padj), padj < 0.05, log2FoldChange > 1) |>
  pull(gene_id) |>
  strip_version()

universe <- res_df |>
  filter(!is.na(padj)) |>
  pull(gene_id) |>
  strip_version()

length(up)
```

## GO over-representation with clusterProfiler

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[Run a GO Biological Process over-representation test on the up-regulated genes with clusterProfiler, using `org.Hs.eg.db`, Ensembl IDs, and the universe as background. Show the top terms.]{.prompt-text}
:::

```{r enrichgo}
#| message: false
#| warning: false
library(clusterProfiler)
library(org.Hs.eg.db)

ego <- enrichGO(
  gene          = up,
  universe      = universe,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.10,
  readable      = TRUE
)

as.data.frame(ego) |>
  dplyr::select(Description, GeneRatio, p.adjust, Count) |>  # AnnotationDbi masks select()
  head(15)
```

## Visualize the enriched terms

::: {.claude-prompt}
[Try this prompt]{.prompt-label}

[Make a dotplot of the top enriched GO terms.]{.prompt-text}
:::

```{r dotplot}
#| fig-height: 6
if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
  dotplot(ego, showCategory = 15) +
    ggplot2::labs(title = "GO Biological Process — up in hypoxia")
} else {
  message("No enriched terms at the chosen thresholds on this subset.")
}
```

::: {.callout-note}
With a subset of reads and only two replicates per group, the exact terms vary, but the **hypoxia / metabolic** signal should dominate. On the full dataset the enrichment is sharper.
:::

## Where to go further

Ask Claude Code to extend the analysis:

- **GSEA** instead of over-representation — `clusterProfiler::gseGO()` on a ranked gene list (uses every gene, no hard cutoff).
- **Pathways** — KEGG (`enrichKEGG`) or Reactome (`ReactomePA::enrichPathway`).
- **MSigDB Hallmark** gene sets via `msigdbr` + `enricher` — the *HALLMARK_HYPOXIA* set is the obvious target here.

→ Recap and how to use this on your own data — **[6 · Wrap-up](06-wrap-up.qmd)**.
